Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-10 demystify.tex
mend
updates

There are various guides on writing SOP, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. In short, focus on your research motivation and goal, and convince us that you can achieve it through your experience, e.g., published papers. You should also back up your claims with concrete evidences. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on project X, then show some results (paper submitted (or even rejected), achieved X\% performance improvement over ..).

@@

\subsection*{Additional Resources}
\begin{itemize}
  \item \href{http://www.pl-enthusiast.net/2022/10/03/how-to-write-a-grad-school-personal-statement/}{How to Write a Grad School Personal Statement} by Mike Hicks
  \item     \href{https://cs-sop.notion.site/cs-sop/CS-PhD-Statements-of-Purpose-df39955313834889b7ac5411c37b958d?p=f5d5980a71524ebaa4e6ae57266b847c&pm=s}{CS PhD SOP database} by cs-sop.org
\end{itemize}
% \subsection{Outline of a LOR}\label{sec:lor-outline}

@@

However, if you have research-based work experiences, then they can be useful and should be mentioned in your SOP and LORs. For example, if you have worked at a research lab (e.g., Microsoft, VinAI in Vietnam, etc), then you should mention that in your SOP and ask your supervisor to write about your research experiences and potentials in their LOR (\autoref{sec:lor}).

@@

\paragraph{Locally} The most common way to find research is right at your own institution.  
If you did well or like a class, you can check with the professor of that class for research opportunities.  
You can also go to the department directory and then professors' websites and see if they are looking for undergraduate researchers.
Even if they say they are looking for graduate students, you still should contact them and ask.

@@

You can also take honor thesis or independent study courses with a professor.  This is a good way to get research experience and also get a LoR from the professor.  You can also ask your academic adviser or other faculty members for suggestions.  Finally, you can also ask your peers who are already doing research.  You can use the method described in \autoref{sec:contact} to contact professors

@@

% \paragraph{Online Research Communities and Open Source Contributions} Places such as Github offer many opportunities for research, even though they might not appear so (e.g., they do not explicitly mention research). 
% In many cases, projects are directly related to research (e.g., open-source software from professors' or research labs). For example, my research group has a many open-source projects that undergrads can contribute to (see \url{https://github.com/dynaroars/}).

%  By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field. Not only you gain research experience, you might be able to get a LoR from the project maintainer and definitely should write about this experience in your SOP (\autoref{sec:research-statement}.)

% \paragraph{Virtual Research Opportunities Beyond Physical Boundaries}: Several universities, organizations, and research institutions offer virtual internships and research programs aimed at providing hands-on research experience. These programs often involve working remotely under the guidance of experienced mentors and collaborating with a team of fellow researchers. For instance, \href{https://docs.google.com/forms/d/1btIwt4HwjyKMOUk-EMy3rbkfWzFxv2lNrMm_zkd0pA4/viewform?edit_requested=true}{UIUC+ Summer Undergraduate Research in Software Engineering}  offers an unpaid remote internship for software engineering students all over the world to collaborate with mentors from University of Illinois at Urbana Champaign.

% \paragraph{Online Conferences and Workshops}: Attending online conferences and workshops is another way to gain exposure to cutting-edge research and establish connections with experts in the field. Many conferences now provide virtual participation options, enabling students to access research talks, poster sessions, and panel discussions and sometimes access designated chat rooms or networking events where participants can engage with researchers, ask questions, and seek potential research collaborations. It is beneficial to also create profiles academic collaboration platforms such as ResearchGate. By creating profiles, joining relevant research groups, and participating in discussions, students can connect with established researchers, contribute to ongoing projects, and potentially collaborate on publications or research proposals as they provide access

>>>>

There are various guides on writing SOP , and \href{https://cs-sop.org/}{many example statements} are available. In short, focus on your research motivation and goal, and convince us that you can achieve it through your experience, e.g., published papers. You should also back up your claims with concrete evidences. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on project X, then show some results (paper submitted (or even rejected), achieved X\% performance improvement over ..).

@@

\subsection*{Additional Resources}
\begin{itemize}
    \item \href{https://chrisblattman.com/blog/2022/01/11/}{Writing your statement of purpose} by Chris Blattman 
  \item \href{http://www.pl-enthusiast.net/2022/10/03/how-to-write-a-grad-school-personal-statement/}{How to Write a Grad School Personal Statement} by Mike Hicks
  \item     \href{https://cs-sop.notion.site/cs-sop/CS-PhD-Statements-of-Purpose-df39955313834889b7ac5411c37b958d?p=f5d5980a71524ebaa4e6ae57266b847c&pm=s}{CS PhD SOP database} by cs-sop.org
\end{itemize}
% \subsection{Outline of a LOR}\label{sec:lor-outline}

@@

However, if you have research-based work experiences, then they can be useful and should be mentioned in your SOP and LORs. For example, if you have worked at a research lab (e.g., Microsoft, DoE, VinAI in Vietnam, etc) (\autoref{sec:research-ability}), then you should mention that in your SOP and ask your supervisor to write about your research experiences and potentials in their LOR (\autoref{sec:lor}).

@@

\paragraph{Locally} Start looking for research opportunities at your own institution.  
If you did well or like a class, you can check with the professor of that class for research opportunities.  
You can also go to the department directory and then professors' websites and see if they are looking for undergraduate researchers.
Even if they say they are looking for graduate students, you still should contact them and ask.

@@

You can also take honor thesis or independent study courses with a professor.  This is a good way to get research experience and also get a LoR from the professor.  You can also ask your academic adviser or other faculty members for suggestions.  Finally, you can also ask your peers who are already doing research.  Use the method described in \autoref{sec:contact} to contact professors

@@

\paragraph{Online Research Communities and Open Source Contributions} Places such as Github offer many opportunities for research, even though they might not appear so (e.g., they do not explicitly mention research). 
In many cases, professors' or research labs put their research projects on Github. For example, my research group has a many open-source projects that undergrads can contribute to (see \url{https://github.com/dynaroars/}).

By contributing code, fixing bugs, implementing new features, or providing documentation, students can gain practical research experience and interact with experienced developers and researchers in the field. Not only you gain research experience, you might be able to get a LoR from the project maintainer and definitely should write about this experience in your SOP (\autoref{sec:research-statement}.)

\paragraph{Virtual Research Opportunities}: This is less common but several places offer virtual internships and research programs aimed at providing hands-on research experience. These programs often involve working remotely under the guidance of experienced mentors and collaborating with a team of fellow researchers. For instance, \href{https://docs.google.com/forms/d/1btIwt4HwjyKMOUk-EMy3rbkfWzFxv2lNrMm_zkd0pA4/viewform?edit_requested=true}{UIUC+ Summer Undergraduate Research in Software Engineering}  offers an unpaid remote internship for software engineering students all over the world to collaborate with mentors from University of Illinois at Urbana Champaign.

% \paragraph{Online Conferences and Workshops}: Attending online conferences and workshops is another way to gain exposure to cutting-edge research and establish connections with experts in the field. Many conferences now provide virtual participation options, enabling students to access research talks, poster sessions, and panel discussions and sometimes access designated chat rooms or networking events where participants can engage with researchers, ask questions, and seek potential research collaborations. By creating profiles, joining relevant research groups, and participating in discussions, students can connect with established researchers, contribute to ongoing projects, and potentially collaborate on publications or research proposals as they provide access
